Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-03-19 paper/manuscript.Rmd
few minor edits to sentences

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
of swimming for quadrupedal mammals is significantly higher than walking or
running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985]. Despite
a large number of coastal and island caribou herds, only a handful of examples
have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012].

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m
per minute). This observation prompted us to investigate whether swimming
behaviour was common among GPS radio-collared caribou in the population (n = 29
adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
Canada). We identified swimming events as two consecutive GPS (2-hour relocation
rates) locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - December 31) and we
identified `r nids` adult female caribou that collectively had `r nevents` swimming events
over three years (Figure 2). In addition to our own visual and remotely sensed
observations, residents of Fogo Island have also reported observing caribou
swimming between islands on numerous occasions.

<!-- add numbers in here -->
In total, `r nevents` swimming events occurred throughout the ice-free season,
with a peak in November. On average, caribou swam approximately `r neventidMean`
(range = `r neventidMin`-`r neventidMax`) per year and remained on those islands
for approximately `r dur` (range = `r durMin`-`r durMax`) days before swimming
again. Caribou may swim between islands in the ocean for similar reasons they
swim in freshwater [@Leblond2016]. We propose that one such explanation, the
forage limitation hypothesis, is a likely explanation for oceanic swimming for
caribou that live on islands in the Fogo Island archipelago.
<!--Is it better
to just state our results here, or give this sort of teaser? Initially I had "86
events by 3 individuals" here, but then we ignore that for a couple paragraphs
of context before getting back into it... which seemed weird
QW: I agree. I've added some Fogo context-->

Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. Caribou were introduced to Fogo Island (n = 26
animals between 1964–67) and Change Island (n = 5 animals in 1964) from
Newfoundland as part of a series of translocations and introductions throughout
the province (Bergerud & Mercer 1989). It is unknown when caribou began
colonizing nearby islands, but given that some are small, over-grazing by newly
arrived caribou may rapidly deplete forage on an annual cycle [@Bergerud1990]. Although
unconfirmed, it is also possible that Fogo Island has experienced forage
depletion since the introduction of caribou in the 1960s. More likely, however,
is fine-scale competition for foraging resources that drives caribou to swim to
new islands.

@@

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
events, suggesting that for ~10% of narjed individuals swimming is relatively
common, on average occurring approximately every `r indMean` days during
ice-free seasons. We did not detect any swimming events for the remaining 17
individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare for most caribou. Overall, there appears to be three
groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
unable to draw conclusions about the prevalence of swimming in the ocean by male
caribou: our initial visual observation was of an adult male, but our GPS data
only included females.

>>>>

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
of swimming for quadrupedal mammals is significantly higher than walking or
running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985]. Despite
the abundance of coastal and island caribou herds, only a handful of examples
have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012].

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m
per minute). This observation prompted us to investigate whether swimming
behaviour was common among GPS radio-collared caribou in the population (n = 29
adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
Canada). We identified swimming events as two consecutive GPS locations (2-hour
relocation rates)  from an individual occurred on different islands.
Northeastern Newfoundland typically experiences pack ice during winter and
caribou are known locally to travel between islands by walking over the ice. We
restricted our GPS data to the ice-free period of the year (April 1 - December
31) and we identified `r nids` adult female caribou that collectively had `r
nevents` swimming events over three years (Figure 2). In addition to our own
visual and remotely sensed observations, residents of Fogo Island have also
reported observing caribou swimming between islands on numerous occasions.

<!-- add numbers in here -->
In total, `r nevents` swimming events occurred throughout the ice-free season,
with a peak in November. On average, caribou swam approximately `r neventidMean`
(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
swimming again. Caribou may swim between islands in the ocean for similar
reasons they swim in freshwater [@Leblond2016]. We propose that one such
explanation, the forage limitation hypothesis, is a likely explanation for
oceanic swimming for caribou that live on islands in the Fogo Island
archipelago.
<!--Is it better
to just state our results here, or give this sort of teaser? Initially I had "86
events by 3 individuals" here, but then we ignore that for a couple paragraphs
of context before getting back into it... which seemed weird
QW: I agree. I've added some Fogo context-->

Forage limitation and over-grazing is a major concern for caribou populations:
it can reduce female reproductive success [@Schaefer2016] and thus could lead to
movement between islands. Forage scarcity has been proposed as a potential
reason for caribou moving between arctic islands on the sea-ice in winter
[@Miller1977]. Caribou were introduced to Fogo Island (n = 26 animals between
1964–67) and Change Island (n = 5 animals in 1964) from Newfoundland as part of
a series of translocations and introductions throughout the province (Bergerud &
Mercer 1989). It is unknown when caribou began colonizing nearby islands, but
given that some are small, over-grazing by newly arrived caribou may rapidly
deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
also possible that Fogo Island has experienced forage depletion since the
introduction of caribou in the 1960s. More likely, however, it is fine-scale
competition for foraging resources that drives caribou to swim to new islands.

@@

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
events, suggesting that for ~10% of collared individuals swimming is relatively
common, on average occurring approximately every `r indMean` days during
ice-free seasons. We did not detect any swimming events for the remaining 17
individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare for most caribou. Overall, there appears to be three
groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
unable to draw conclusions about the prevalence of swimming in the ocean by male
caribou: our initial visual observation was of an adult male, but our GPS data
only included females.
